Distance Sampling in \proglang{R}

### 2015-11-22 paper.Rmd
Non-ascii characters replaced
Long dashes for 103-108 and 151-153 replaced.

For both points and lines, once the geometry of the sampler _<likely spell out geom of sampler>_ has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler. For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability. For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.

@@

The line transect data is simulated data based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*). The data is simulated from models fitted to data from the International Whaling Commission’s International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. They consist of 99 observations and include information on the effort expended and whether observations were in one of two geographical strata (near or far from land).

@@

It is important that the detection function accurately models detectability near zero distance. We are less worried by its behaviour further away from 0. To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*). Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103–108, 151–153].

@@

There are many factors that can affect the probability of detecting an object. These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few). We assume that these variables affect detection only via the scale of the detection function (and that they don't affect the shape).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where “given values” are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

>>>>

For both points and lines, once the geometry of the sampler _likely spell out geom of sampler, or note it specification is forthcoming_ has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler. For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability. For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.

@@

The line transect data is simulated data based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*). The data is simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. They consist of 99 observations and include information on the effort expended and whether observations were in one of two geographical strata (near or far from land).

@@

It is important that the detection function accurately models detectability near zero distance. We are less worried by its behaviour further away from 0. To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*). Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103-108, 151-153].

@@

There are many factors that can affect the probability of detecting an object. These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few). We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where 'given values' are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.
